---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(colourlookup)

library(tidyr)
library(ggbeeswarm)
library(ggplot2)
```

# colourlookup

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`{colourlookup}` is a proof-of-concept technical demonstration of an order-preserving
minimal perfect hash in application to R colour handling.

Using a hash to match colour names can be of the order of ~100x faster than 
a linear search.  

### What's in the box

* `col_to_rgb()` a version of `col2rgb()` which uses hash lookup rather than 
  linear search to match R colour names to RGBA values.
#### Benchmark

This is the results of a benchmark comparing R's `col2rgb()` (using linear name search) with this package's
`col_to_rgb()` using a hashed name lookup.

See end of this page for benchmark code.

```{r echo = FALSE, fig.height = 3, fig.width = 8}
cols <- colours()

res <- bench::mark(
  `hash lookup` = col_to_rgb(cols),
  `linear lookup` = col2rgb(cols, alpha = TRUE),
  check = TRUE
)

knitr::kable(res[,1:5])

res$expression <- as.character(res$expression)
ggplot(res) + 
  geom_col(aes(expression, `itr/sec`, fill = expression)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_fill_manual(values = c('dodgerblue4', 'grey50')) + 
  labs(
    x = NULL,
    y = "Iterations / second",
    title = "Speed comparison of hash lookup vs built-in linear search",
    subtitle = "when finding RGBA values for all named colours"
  )

```


Order-preserving minimal perfect hash
===============================================================================

### Hash

A **hash** of a string is just a combination of the bits in the string to 
produce another value.

Useful hashes are those that produce markedly different output for slightly
different inputs. e.g. CRC32, SHA1, MD5

The `hash()` function in `{rlang}` is a good example of such a hash. Note 
how different the output of the hash is when only one character has changed.

```{r}
rlang::hash("hello")
rlang::hash("hello!")
```

### Hash collisions

Due to the limits of reality, hashing arbitrary length strings into shorter strings
will mean that for some special cases different strings will produce the 
same hashed output (see [the pigeonhole principle](https://en.wikipedia.org/wiki/Pigeonhole_principle)).  This is known as a
[hash collision](https://en.wikipedia.org/wiki/Hash_collision)

Hash collisions mean that you don't know for sure which string produced a particular 
hash, and so extra context information is kept to help disambiguate when this occurs.

```
hash("apple")  => 1034  # Collision with "carrot"
hash("banana") => 245
hash("carrot") => 1034  # Collision with "apple"
```



### Perfect Hashing

[Perfect hashing](https://en.wikipedia.org/wiki/Perfect_hash_function) is 
when the input data is known and fixed, and there exists a hash that 
has no collisions.


```
hash("apple")  => 1034 
hash("banana") => 245
hash("carrot") => 876  
```



### Minimal Perfect Hashing

[Minimal Perfect Hashing](https://en.wikipedia.org/wiki/Perfect_hash_function#Minimal_perfect_hash_function) is when a perfect hash can map *n* strings onto *n* consecutive integers. 

Note: the order of the integers does not have to match the original order of the strings.


```
hash("apple")  => 3 
hash("banana") => 1
hash("carrot") => 2  
```

### Order-Preserving Minimal Perfect Hashing

[Order-Preserving minimal perfect hashing](https://en.wikipedia.org/wiki/Perfect_hash_function#Order_preservation) is when
the *n* strings match to *n* consecutive integers in the same order as the strings were
defined.

As an example, if we have 3 strings "apple", "banana" and "carrot", an *order-preserving
minimal perfect hash* would have the following result:

```
hash("apple")  => 1
hash("banana") => 2
hash("carrot") => 3
```


## R colours

R has 657 built-in colour names it understands e.g. "red", "hotpink".  This is 
a fixed list of colours that has not changed membership since originally introduced
(as far as I know).

When looking up colours with `col2rgb()`, R does a linear search over these 657 colours
to find the values to return.

Since the list of colours is fixed and known, this colour lookup could be
replaced by a hash lookup. 

For this package I have used an *order-preserving minimal perfect hash* to do this
lookup.

## Creating the C code for the minimal perfect hash

There are a number of programs which will produce C code to
do minimal perfect hashing of a list of strings e.g. `gperf`, `cmph`.

I chose `mph` as described in [A Family of Perfect Hashing Methods(pdf)](https://staff.itee.uq.edu.au/havas/TR0242.pdf)

* Download and compile [http://www.ibiblio.org/pub/Linux/devel/lang/c/mph-1.2.tar.gz]
* In R: `writeLines(colors(), "input.txt")`
* On the command line:  `mph < input.txt | emitc > hash.c`
* Incorporate the `hash.c` code into this package and add the `col_to_rgb()` wrapper
  which uses it.

## Installation

You can install from [GitHub](https://github.com/coolbutuseless/colourlookup) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/colourlookup')
```

## Example

```{r example}
library(colourlookup)

col_to_rgb(c("hotpink", "#abc", "#aabb99", "#aabb9980"))
```



Benchmarks
===============================================================================

Looking up RGBA values for all named colours in R
-------------------------------------------------------------------------------

```{r eval = FALSE}
cols <- colours()

res <- bench::mark(
  col_to_rgb(cols),
  col2rgb(cols, alpha = TRUE),
  check = TRUE
)
```


```{r echo=FALSE}
knitr::kable(res[,1:5])
plot(res) + theme_bw()
```




## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
