---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(colourlookup)

library(tidyr)
library(ggbeeswarm)
library(ggplot2)
library(farver)
```

# colourlookup

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/colourlookup/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/colourlookup/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{colourlookup}` is a proof-of-concept technical demonstration of an order-preserving
minimal perfect hash in application to R colour handling.

Using a hash to match colour names can be of the order of ~100x faster than 
a linear search.  

### R colours

R has 657 built-in colour names it understands e.g. "red", "hotpink".  This is 
a fixed list of colours that has not changed membership since originally introduced
(as far as I know).

When looking up colours with base R `col2rgb()`, a linear search over these 657 colours
is performed to find the matching RGBA values.

Since the list of colours is fixed and known, this colour lookup can be
replaced by a hash lookup which is faster.

In particular, this package uses an *order-preserving minimal perfect hash* - 
see the technical details at the bottom of this page.


### What's in the box

* `col_to_rgb()` a version of `col2rgb()` which uses hash lookup rather than 
  linear search to match R colour names to RGBA values.
#### Benchmark

### Benchmark 

This benchmark compares named colour lookup of RGBA values with:

* baseR's `col2rgb()` (using linear name search)
* colour lookup using `farver::decode_colour()`
* this package's `col_to_rgb()` using a hashed name lookup.


<details>
<summary>Click to show/hide benchmarking code</summary>

```{r echo = TRUE}
cols <- sample(colours(), 10000, T)

res <- bench::mark(
  `this package (hash)` = col_to_rgb(cols),
  `farver` = decode_colour(cols),
  `baseR (linear)` = col2rgb(cols, alpha = TRUE),
  check = FALSE
)
knitr::kable(res[,1:5])
```
</details>


```{r echo = FALSE, fig.height = 3, fig.width = 8}
res$expression <- as.character(res$expression)
ggplot(res) + 
  geom_col(aes(expression, `itr/sec`, fill = expression)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_fill_manual(values = c('grey50', 'grey50', 'dodgerblue4')) + 
  labs(
    x = NULL,
    y = "Iterations / second",
    title = "Speed comparison of hash lookup vs built-in linear search",
    subtitle = "Finding RGBA values for 10,000 named colours"
  )

```


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/colourlookup) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/colourlookup')
```

## Example

```{r example}
library(colourlookup)

col_to_rgb(c("hotpink", "#abc", "#aabb99", "#aabb9980"))
```


## Creating the C code for the minimal perfect hash

There are a number of programs which will produce C code to
do minimal perfect hashing of a list of strings e.g. `gperf`, `cmph`.

I chose `mph` as described in [A Family of Perfect Hashing Methods(pdf)](https://staff.itee.uq.edu.au/havas/TR0242.pdf)

* Download and compile [http://www.ibiblio.org/pub/Linux/devel/lang/c/mph-1.2.tar.gz]
* In R: `writeLines(colors(), "input.txt")`
* On the command line:  `mph < input.txt | emitc > hash.c`
* Incorporate the `hash.c` code into this package and add the `col_to_rgb()` wrapper
  which uses it.



Benchmarks
===============================================================================

Looking up RGBA values for all named colours in R
-------------------------------------------------------------------------------

Example

```{r}
col_to_rgb(c('white', 'blue'))
```

```{r echo=FALSE}
set.seed(1)
```

<details>
<summary>Click to show/hide benchmarking code</summary>

```{r eval = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Lots of named colours
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colours(), 10000, T)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Check we all agree on the RGB representation
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
a1 <- col_to_rgb(cols)
a2 <- col2rgb(cols, alpha = TRUE)
a3 <- farver::decode_colour(cols) |> t()
a3 <- rbind(a3, 255L)
rownames(a3) <- c('red', 'green', 'blue', 'alpha')
identical(a1, a2)
identical(a2, a3)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Benchmark
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(farver)
res <- bench::mark(
  colourlookup = col_to_rgb(cols),
  baseR        = col2rgb(cols, alpha = TRUE),
  farver       = decode_colour(cols, alpha = TRUE),
  check = FALSE
)
knitr::kable(res[,1:5])
```

</details>

```{r echo=FALSE}
plot(res) + theme_bw()
```


Looking up RGBA values for hex colours in R
-------------------------------------------------------------------------------

Example

```{r}
col_to_rgb(c('#12345678', '#ab1276'))
```


<details>
<summary>Click to show/hide benchmarking code</summary>


```{r eval = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Lots of hex colours
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colours(), 10000, T) |>
  col_to_rgb() |>
  t() |>
  rgb(maxColorValue = 255)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Check we all agree on the RGB representation
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
a1 <- col_to_rgb(cols)
a2 <- col2rgb(cols, alpha = TRUE)
a3 <- farver::decode_colour(cols) |> t()
a3 <- rbind(a3, 255L)
rownames(a3) <- c('red', 'green', 'blue', 'alpha')
identical(a1, a2)
identical(a2, a3)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# benchmark
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(farver)
res <- bench::mark(
  colourlookup = col_to_rgb(cols),
  baseR        = col2rgb(cols, alpha = TRUE),
  farver       = decode_colour(cols, alpha = TRUE),
  check = FALSE
)
knitr::kable(res[,1:5])
```

</details>


```{r echo=FALSE}
plot(res) + theme_bw()
```



Looking up native raster "packed integer" for hex colours in R
-------------------------------------------------------------------------------

Example

```{r}
col_to_packed_int(c('#12345678', '#ab1276'))
```


<details>
<summary>Click to show/hide benchmarking code</summary>

```{r eval = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Lots of hex colours
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colours(), 10000, T) |>
  col_to_rgb() |>
  t() |>
  rgb(maxColorValue = 255)


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# benchmark
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(farver)
res <- bench::mark(
  `this package` = col_to_packed_int(cols),
  farver         = encode_native(cols),
  check = FALSE
)

knitr::kable(res[,1:5])
```

</details>

```{r echo=FALSE}
plot(res) + theme_bw()
```



Technical Details: Order-preserving minimal perfect hash
===============================================================================

### Hash

A **hash** of a string is just a combination of the bits in the string to 
produce another value.

Useful hashes are those that produce markedly different output for slightly
different inputs. e.g. CRC32, SHA1, MD5

The `hash()` function in `{rlang}` is a good example of such a hash. Note 
how different the output of the hash is when only one character has changed.

```{r}
rlang::hash("hello")
rlang::hash("hello!")
```

### Hash collisions

Due to the limits of reality, hashing arbitrary length strings into shorter strings
will mean that for some special cases different strings will produce the 
same hashed output (see [the pigeonhole principle](https://en.wikipedia.org/wiki/Pigeonhole_principle)).  This is known as a
[hash collision](https://en.wikipedia.org/wiki/Hash_collision)

Hash collisions mean that you don't know for sure which string produced a particular 
hash, and so extra context information is kept to help disambiguate when this occurs.

```
poor_hash("apple")  => 1034  # Collision with "carrot"
poor_hash("banana") => 245
poor_hash("carrot") => 1034  # Collision with "apple"
```



### Perfect Hashing

[Perfect hashing](https://en.wikipedia.org/wiki/Perfect_hash_function) is 
when the input data is known and fixed, and there exists a hash that 
has no collisions.


```
perfect_hash("apple")  => 1034 
perfect_hash("banana") => 245
perfect_hash("carrot") => 876  
```



### Minimal Perfect Hashing

[Minimal Perfect Hashing](https://en.wikipedia.org/wiki/Perfect_hash_function#Minimal_perfect_hash_function) is when a perfect hash can map *n* strings onto *n* consecutive integers. 

Note: the order of the integers does not have to match the original order of the strings.


```
min_perfect_hash("apple")  => 3 
min_perfect_hash("banana") => 1
min_perfect_hash("carrot") => 2  
```

### Order-Preserving Minimal Perfect Hashing

[Order-Preserving minimal perfect hashing](https://en.wikipedia.org/wiki/Perfect_hash_function#Order_preservation) is when
the *n* strings match to *n* consecutive integers in the same order as the strings were
defined.

As an example, if we have 3 strings "apple", "banana" and "carrot", an *order-preserving
minimal perfect hash* would have the following result:

```
omph_hash("apple")  => 1
omph_hash("banana") => 2
omph_hash("carrot") => 3
```
