---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(colorfast)

library(tidyr)
library(ggbeeswarm)
library(ggplot2)
library(farver)
```

# colorfast

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/colorfast/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/colorfast/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{colorfast}` is a package for converting R colors and hexadecimal colors
to arrays of RGBA values, and packed integer format (suitable for native rasters).


### What's in the box

* `col_to_rgb()` a faster version of `col2rgb()` 
* `col_to_int()` for conversion of colors to packed integer format used 
  in native rasters (4 bytes for each RGBA color channel are packed into a
  single R integer)


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/colorfast) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/colorfast')
```

## Example

```{r example}
library(colorfast)

col_to_rgb(c("hotpink", "#abc", "#aabb99", "#aabb9980"))
col_to_int(c("hotpink", "#abc", "#aabb99", "#aabb9980"))
```


## Benchmarks

The following benchmarks compare `{colorfast}` to `{farver}` and base R.

| package       | Convert to RGBA array | Convert to packed integer |
|---------------|-----------------------|---------------------------|
| Base R        | `col2rgb()`           | N/A                       |
| `{farver}`    | `decode_colour()`     | `encode_native()`         |
| `{colorfast}` | `col_to_rgb()`        | `col_to_int()`            |


<details>
<summary>Click to show raw benchmark results</summary>

```{r echo = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# R color names
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colors(), 10000, T)


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Check we all agree on the RGB representation
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
a1 <- col_to_rgb(cols)
a2 <- col2rgb(cols, alpha = TRUE)
a3 <- farver::decode_colour(cols) |> t()
a3 <- rbind(a3, 255L)
rownames(a3) <- c('red', 'green', 'blue', 'alpha')
stopifnot(identical(a1, a2))
stopifnot(identical(a2, a3))


res_col <- bench::mark(
  `baseR         ` = col2rgb(cols, alpha = TRUE),
  `{farver}`       = decode_colour(cols),
  `{colorfast}`    = col_to_rgb(cols),
  check = FALSE
)
knitr::kable(res_col[,1:5])



#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#
# Hex colors
#
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colors(), 100000, T) |>
  col_to_rgb() |>
  t() |>
  rgb(maxColorValue = 255)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Check we all agree on the RGB representation
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
a1 <- col_to_rgb(cols)
a2 <- col2rgb(cols, alpha = TRUE)
a3 <- farver::decode_colour(cols) |> t()
a3 <- rbind(a3, 255L)
rownames(a3) <- c('red', 'green', 'blue', 'alpha')
stopifnot(identical(a1, a2))
stopifnot(identical(a2, a3))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# benchmark
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(farver)
res_hex <- bench::mark(
  baseR         = col2rgb(cols, alpha = TRUE),
  `{farver}`    = decode_colour(cols, alpha = TRUE),
  `{colorfast}` = col_to_rgb(cols),
  check = FALSE
)
knitr::kable(res_hex[,1:5])



#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Conversion to packed int
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cols <- sample(colors(), 100000, T) |>
  col_to_rgb() |>
  t() |>
  rgb(maxColorValue = 255)


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# benchmark
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(farver)
res_int <- bench::mark(
  `{farver}`    = encode_native(cols),
  `{colorfast}` = col_to_int(cols),
  check = TRUE
)

knitr::kable(res_int[,1:5])
```
</details>



```{r echo = FALSE, fig.height = 3, fig.width = 8}
library(ggplot2)
res_col$expression <- as.character(res_col$expression)
ggplot(res_col) + 
  geom_col(aes(expression, `itr/sec`, fill = expression)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_fill_manual(values = c('dodgerblue4', 'grey50', 'grey50')) + 
  labs(
    x = NULL,
    y = "Iterations / second",
    title = "Speed comparison: {colorfast} vs {farver} vs base",
    subtitle = "Converting 10,000 named colors to RGBA values"
  )
```


```{r echo = FALSE, fig.height = 3, fig.width = 8}
library(ggplot2)
res_hex$expression <- as.character(res_hex$expression)
ggplot(res_hex) + 
  geom_col(aes(expression, `itr/sec`, fill = expression)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_fill_manual(values = c('dodgerblue4', 'grey50', 'grey50')) + 
  labs(
    x = NULL,
    y = "Iterations / second",
    title = "Speed comparison: {colorfast} vs {farver} vs base",
    subtitle = "Converting 100,000 hexadecimal colors to RGBA values"
  )
```


```{r echo = FALSE, fig.height = 3, fig.width = 8}
library(ggplot2)
res_int$expression <- as.character(res_int$expression)
ggplot(res_int) + 
  geom_col(aes(expression, `itr/sec`, fill = expression)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  scale_fill_manual(values = c('dodgerblue4', 'grey50', 'grey50')) + 
  labs(
    x = NULL,
    y = "Iterations / second",
    title = "Speed comparison: {colorfast} vs {farver}",
    subtitle = "Converting 100,000 hexadecimal colors to RGBA values"
  )
```




